Template for manuscript

### 2024-12-09 LakeMetabolism_project/Synthesis_21_22_23/report/manuscript.Rnw
Work on manuscript

\documentclass[fleqn,11pt,lineno]{manuscript}
%%\usepackage{setspace}
%%\doublespacing
\usepackage{soul}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=blue,      
    urlcolor=blue,
    citecolor=blue,
    }

@@

In 2021, 2022, and 2023, all 12 lakes were monitored over several days.
For that purpose, EXO2 multiparameter sondes were installed
(YSI, Yellow Springs, OH, USA), tracking ecosystem parameters with high
frequency (2-minute intervals in 2021 and 2022, 5-minute intervals in
2023 with the exception of ERL122, which was monitored in 15-minute intervals).
For the purpose of this study, only dissolved oxygen and temperature measurements
yielded from these sondes are relevant. The sensors were situated at a water depth
of approximately 1-1.5 m in each lake. All optical sensors were
wiped clean before every measurement with a built-in wiper.
The monitoring period was 16 September-24 September in 2021, 22 June-3 July in 2022,
and 22 June-17 July in 2023.

@@

Dissolved oxygen and water temperature measurements were yielded from EXO2 multiparameter
sondes, as described in section~\nameref{subsec:design}. For the purpose
of estimating ecosystem metabolism, wind and irradiation data were yielded
from the Programme for Monitoring of the Greenland Ice Sheet (PROMICE), providing automatic
weather station data \citep{Fausto:2021}.
The dataset named \texttt{QAS\_L\_hour.csv} originating from the QAS\_L automated weather station
near Narsarsuaq was downloaded from a
\href{https://dataverse.geus.dk/citation?persistentId=doi:10.22008/FK2/IW73UU}{public database}
provided by PROMICE \citep{metadata}.

\subsection{Data preparation}

The data preparation for estimating lake metabolism was done separately for each monitoring
period (16 September-24 September 2021, 22 June-3 July 2022, and 22 June-17 July 2023).

\subsubsection*{Sonde data}

First, the raw sonde data from each lake were imported and merged into
one dataset for a given monitoring period. The variables needed for lake metabolism
estimation were water temperature (\textdegree{}C) and
dissolved oxygen (mg/L). Next, an outlier analysis was performed by removing
values higher than three times the median absolute deviation of all values in a
sliding window of one day window size \citep{Luerig:2021}.
After outlier removal, the time series of water temperature and dissolved oxygen
were subsequently investigated for larger data gaps due to potential lack of sensor
measurements over certain periods.

To ensure overlapping estimation of lake metabolism, the data was subsequently
cut to achieve equal starting and ending time points for all lakes during a given
observation period. Due to large data gaps during the period of 2022, we chose
to keep as much complete days as possible in order to estimate the maximum
amount of days, rather than selecting the longest common monitoring period
for all lakes.

After cutting, missing values were replaced by a
weighted moving average, where the weights of observations around the
central value decrease exponentially \citep{Moritz:2017}.
For larger data gaps spanning over several hours, we chose
to not estimate lake metabolism for the affected days, rather than imputing these values.

\subsubsection*{Weather data}

\subsection{Statistical analysis}

\subsection{Implementation}

>>>>

\documentclass[11pt,lineno]{manuscript}
%\setlength{\mathindent}{0pt}
%%\usepackage{setspace}
%%\doublespacing
\usepackage{soul}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=blue,      
    urlcolor=blue,
    citecolor=blue,
    }

@@

In 2021, 2022, and 2023, all 12 lakes were monitored over several days.
For that purpose, EXO2 multiparameter sondes were installed
(YSI, Yellow Springs, OH, USA), tracking ecosystem parameters with high
frequency (2-minute intervals in 2021 and 2022, 5-minute intervals in
2023 with the exception of ERL122, which was monitored in 15-minute intervals).
For the purpose of this study, only dissolved oxygen (hereafter DO) and temperature measurements
yielded from these sondes are relevant. The sensors were situated at a water depth
of approximately 1-1.5 m in each lake. All optical sensors were
wiped clean before every measurement with a built-in wiper.
The monitoring period was 16 September-24 September in 2021, 22 June-3 July in 2022,
and 22 June-17 July in 2023.

@@

DO ($mg\,L^{-1}$) and water temperature (\textdegree{}$C$) measurements
were yielded from EXO2 multiparameter
sondes, as described in section~\nameref{subsec:design}. For the purpose
of estimating ecosystem metabolism, wind and irradiation data were yielded
from the Programme for Monitoring of the Greenland Ice Sheet (PROMICE), providing automatic
weather station data \citep{Fausto:2021}.
The dataset named \texttt{QAS\_L\_hour.csv} originating from the QAS\_L automated weather station
near Narsarsuaq was downloaded from a
\href{https://dataverse.geus.dk/citation?persistentId=doi:10.22008/FK2/IW73UU}{public database}
provided by PROMICE \citep{metadata}.
Wind data was provided as average wind speed in hourly intervals ($m\,s^{-1}$), while
irradiation was provided as downwelling shortwave irradiation additionally
corrected for the tilt of the sensor ($W\,m^{-2}$).

\subsection{Data analysis pipeline} \label{subsec:pipeline}

The data preparation and analysis for estimating lake metabolism was done separately for each monitoring
period (16 September-24 September 2021, 22 June-3 July 2022, and 22 June-17 July 2023).

First, the raw sonde data from each lake were imported and merged into
one dataset for a given monitoring period. The variables required for lake metabolism
estimation were water temperature and
DO. Next, an outlier analysis was performed by removing
values higher than three times the median absolute deviation of all values in a
sliding window of one day window size \citep{Luerig:2021}.
After outlier removal, the time series of water temperature and DO
were plotted and subsequently investigated for larger data gaps due to potential lack of sensor
measurements over certain periods.

To ensure overlapping estimation of lake metabolism, the data was subsequently
cut to achieve equal starting and ending time points for all lakes during a given
monitoring period. Due to large data gaps during the period of 2022, we chose
to keep as much complete days as possible in order to estimate the maximum
amount of days, rather than selecting the longest common monitoring period
for all lakes.

After cutting the sonde data, they were subsequently merged with the weather data.
Since the weather data were collected in hourly intervals, they were inflated to match
the measurement intervals of the sonde data (see section \nameref{subsec:design}).

After merging sonde and weather data, potential missing values due to outlier removal
were imputed by a weighted moving average, where the weights of observations around the
central value to be imputed decrease exponentially \citep{Moritz:2017}.
For larger data gaps spanning over several hours (due to sonde malfunction rather
than outlier removal), we chose
to not estimate lake metabolism for the affected days.\smallskip

All data preparation and analysis was done with the \texttt{R} programming language \citep{R}.
To estimate lake metabolism, the package \texttt{LakeMetabolizer} was used \citep{Winslow:2016}.
Specifically, a Bayesian framework to estimate whole lake metabolism from free-water
DO by using the function \texttt{metab.bayesian} was chosen for
this analysis \citep{Holtgrieve:2010}.
In order to be able to fit the model, a minimum of 6 variables are needed as input:

\begin{enumerate}
\item doobs: DO concentration measurements ($mg\,L^{-1}$)
\item do.sat: Equilibrium DO concentration for specific temperature ($mg\,L^{-1}$, hereafter do.sat)
\item k.gas: Gas and temperature specific gas transfer coefficient ($m^{-1}$, hereafter k.gas)
\item par: Photosynthetically active radiation ($\mu{}mol\,m^{-2}\,s^{-1}$, hereafter par)
\item z.mix: Actively mixed layer depth ($m$, hereafter z.mix)
\item wtr: Water temperature (\textdegree{}$C$)
\end{enumerate}

Some of the variables listed above have to be derived from other time series data prior
to estimating lake metabolism,
namely do.sat, k.gas, par, and z.mix.
do.sat was derived by using the function \texttt{o2.at.sat}, which calculates
the equilibrium DO concentration of oxygen in water at supplied conditions.
To estimate k.gas, a gas transfer coefficient model must be fit.
The \texttt{LakeMetabolizer} package offers different methods, and we chose
an empirical wind-based gas exchange model, introduced by Vachon and
Prairie in 2013 \citep{Vachon:2013}. This model not only takes wind into
consideration, but also lake area.
All models of gas exchange 
return a k\textsubscript{600} value, a gas exchange normalized to a Schmidt 
number (Sc) of 600, or O\textsubscript{2} at 17.5 \textdegree{}$C$.
After estimating k\textsubscript{600}, it must be transformed to k.gas by using the
function \texttt{k600.2.kGAS} provided by the \texttt{LakeMetabolizer}
package. par was derived from shortwave irradiation,
which we had available (see section \nameref{subsec:source}).
To achieve this transformation, we used the function \texttt{sw.to.par},
which uses the simple empirical transformation $par = 2.114 \times sw$ \citep{Britton:1976}.
To estimate z.mix, one normally needs a time series profile of water temperature
or water density. Since we did not have such data available, we decided to
simulate actively mixed layer depth by drawing from a uniform distribution
with limits 0 and maximum depth of the lake. The reasoning behind this simulation
is that, although the simulated time series of z.mix likely differing substantially
from the actual, a possible treatment effect of fish versus no fish should still be
consistent, given the simulation is performed equally in all lakes.\smallskip

After performing the data preparation and the derivation for certain model
inputs, the Bayesian lake metabolism model was fitted for each lake and
monitoring period. The Bayesian model is special, as it considers 2 types
of errors: Differences between the true data generating process and the process
defined in the model (process error) and errors resulting from inaccuracies
of DO measurements (observation error). Reflecting this philosophy, the
Bayesian model distinguishes between 3 categories of DO values: $y$ represents
the DO measurements, $\alpha$ the true but unknown DO values, and
$\alpha^{\ast}$ the model's estimates of the true value.
The observed DO measurements $y$ are modeled as random deviations from the true
value $\alpha$:

\begin{equation}
y_{t} \sim N(\alpha_{t}, \tau_{v}),
\label{eq:doobs}
\end{equation}

where $\tau_{v}$ is the precision of the observation error (precision is
the reciprocal of variance).
Note, that the Bayesian model fits parameter values by making a comparison
between observed and theoretical values of DO.
This comparison is made in equation \ref{eq:dotrue}, where we define the true value of 
DO at time $t$, $\alpha_{t}$, as being normally distributed with mean 
$\alpha^{\ast}_{t}$ and process precision $\tau_{w}$:

\begin{equation}
\alpha_{t} \sim N(\alpha^{\ast}_{t}, \tau_{w}),
\label{eq:dotrue}
\end{equation}

where $\tau_{w}$ is the precision of the process error.
The important distinction of process and observation error is that
process error affects the state of the system at time
$t + k(k=0, 1, ..., T-t)$, because the state evolves dynamically, whereas
observation error only affects the state of the system at time $t$.

Both $\tau_{v}$ and $\tau_{w}$ are given minimally informative priors
following a gamma distribution with shape and rate parameters of 0.001.

The Bayesian model can be described by the following equations:

\begin{gather}
x^{2} \label{eq:sys1}\\
x^{3} \label{eq:sys2}
\end{gather}
